Two subtle issues with over-representation analysis: Demonstration

### 2024-07-01 app/report.Rmd
FDR fix

Source: https://github.com/markziemann/background

@@

### Genes without annotated categories are discarded

@@

### p-value adjustment is applied incorrectly

@@

```{r,checklibraries}

@@

This software was developed by Mark Ziemann, report issues at the [GitHub repository](https://github.com/markziemann/background).

@@

```{r, gsl1}

genesetlibrary

nsets <- length(unique(mygs$term))
message(paste("Number of sets:",nsets))
```

@@

```{r,gsl2}
summary(table(mygs$term))
```

@@

  ora$ES <- gr/br
  ora$ID = ora$geneID = ora$p.adjust = ora$Count = NULL
  colnames(ora) <- gsub("qvalue","FDR",colnames(ora))
  colnames(ora) <- gsub("GeneRatio","FgRatio",colnames(ora))
  ora <- ora[,c("Description","FgRatio","BgRatio","ES","pvalue","FDR")]
  ora$ES <- signif(ora$ES,3)
  ora$pvalue <- signif(ora$pvalue,3)
  ora$FDR <- signif(ora$FDR,3)
}

@@

Euler diagram.

@@

if ( comparison == "Background error") {
  v1 <- list("Original"=ora1_sets, "BG fix"=ora2_sets)
}
plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2))

@@

plot_ly(
  m, x = ~ES.x, y = ~ES.y
) %>%
  add_trace(m, x = ~Original, y = ~BGcorrected, type = "scatter",
            mode="lines", line=list(color='grey')) %>%
  add_trace(m, x = ~ES.x, y = ~ES.y, type = "scatter", mode = "markers",
            showlegend = FALSE, text = m$Description,
            hoverinfo = 'text') %>%
  layout(title="Enrichment score comparison")
}

@@

}

>>>>

This software was developed by Mark Ziemann under an MIT licence.
Report issues at the [GitHub repository](https://github.com/markziemann/background).

@@

### Background problem: Genes without annotated categories are discarded

@@

### FDR problem: p-value adjustment is applied incorrectly

@@

@@

The problem we are characterising is:

```{r,notes}

@@

```{r, gsl1}
genesetlibrary
```

Number of sets.

```{r,gsl1a}
length(unique(mygs$term))
```

@@

```{r,gsl2}
summary(as.numeric(unname(table(mygs$term))))
```

@@

  ora$ES <- gr/br
  ora$ID = ora$geneID = ora$p.adjust = ora$Count = NULL
  colnames(ora) <- gsub("qvalue","FDR",colnames(ora))
  colnames(ora) <- gsub("GeneRatio","FgRatio",colnames(ora))
  ora <- ora[,c("Description","FgRatio","BgRatio","ES","pvalue","FDR")]
  ora$ES <- signif(ora$ES,3)
  ora$pvalue <- signif(ora$pvalue,3)
  ora$FDR <- signif(ora$FDR,3)
} else if (comparison == "FDR error") {
  gs1 <- mygs
  gs1 <- gs1[which(gs1$gene %in% bg),]
  terms <- names(which(table(gs1$term)>5))
  gs1 <- gs1[gs1$term %in% terms,]
  options(enrichment_force_universe = FALSE)
  ora <- as.data.frame(enricher(gene = fg ,
                                universe = bg, minGSSize = 5, maxGSSize = 500000, TERM2GENE = gs1,
                                pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

  ora$ES <- gr/br
  ora$ID = ora$geneID = ora$p.adjust = ora$Count = NULL
  colnames(ora) <- gsub("qvalue","FDR",colnames(ora))
  colnames(ora) <- gsub("GeneRatio","FgRatio",colnames(ora))
  ora <- ora[,c("Description","FgRatio","BgRatio","ES","pvalue","FDR")]

  # proper FDR correction
  nsets <- length(which(table(gs1$term)>5))
  nres <- nrow(ora)
  diff <- nsets - nres
  pvals <- c(ora$pvalue,rep(1,diff))
  ora$FDR <- p.adjust(pvals,method="fdr")[1:nrow(ora)]
}

@@

Euler diagram of pathways with FDR<0.05.

@@

if ( comparison == "Background error") {
  v1 <- list("Original"=ora1_sets, "BG fix"=ora2_sets)
} else if (comparison == "FDR error") {
  v1 <- list("Original"=ora1_sets, "FDR fix"=ora2_sets)
}
plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2))

Jaccard similarity index of pathways with FDR<0.05.

```{r,jac1}

jac <- length(intersect(ora1_sets,ora2_sets)) / length(union(ora1_sets,ora2_sets))

message(paste("Jaccard index:",signif(jac,3)))

@@

  plot_ly(
    m, x = ~ES.x, y = ~ES.y
  ) %>%
    add_trace(m, x = ~Original, y = ~BGcorrected, type = "scatter",
              mode="lines", line=list(color='grey')) %>%
    add_trace(m, x = ~ES.x, y = ~ES.y, type = "scatter", mode = "markers",
              showlegend = FALSE, text = m$Description,
              hoverinfo = 'text') %>%
    layout(title="Enrichment score comparison")
} else if (comparison == "FDR error") {
  m$Original <- m$FDRcorrected <- round(VEC[1:nrow(m)],0.1)
  m$ES.x <- signif(m$ES.x,3)
  m$ES.y <- signif(m$ES.y,3)

  plot_ly(
    m, x = ~ES.x, y = ~ES.y
  ) %>%
    add_trace(m, x = ~Original, y = ~FDRcorrected, type = "scatter",
              mode="lines", line=list(color='grey')) %>%
    add_trace(m, x = ~ES.x, y = ~ES.y, type = "scatter", mode = "markers",
              showlegend = FALSE, text = m$Description,
              hoverinfo = 'text') %>%
    layout(title="Enrichment score comparison")
}

@@

} else if (comparison == "FDR error") {
  m$Original <- m$FDRcorrected <- round(VEC[1:nrow(m)],0.1)
  m$logFDR.x <- signif(m$logFDR.x,3)
  m$logFDR.y <- signif(m$logFDR.y,3)

  plot_ly(
    m, x = ~logFDR.x, y = ~logFDR.y
  ) %>%
    add_trace(m, x = ~Original, y = ~FDRcorrected, type = "scatter",
              mode="lines", line=list(color='grey')) %>%
    add_trace(m, x = ~logFDR.x, y = ~logFDR.y, type = "scatter", mode = "markers",
              showlegend = FALSE, text = m$Description,
              hoverinfo = 'text') %>%
    layout(title="log FDR comparison")
}
